Dish: Distribution-oblivious Shell Scripting

### 2019-11-28 pldi/draft.tex
Transformations

\subsection{Graph transformations}
\label{ir:transformations}

Before describing the different types of parallelization
optimizations, we formalize the intuition about stateless and pure
commands that was described in \Cref{cmd}.
%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }
%
As mentioned there, stateless commands operate independently on
individual elements of the stream (characters, lines, or files)
without maintaining any state. Some examples include \ttt{grep},
\ttt{basename}, and \ttt{tr}. Formally, a commmand $f$ is stateless if
it is ``linear'' with respect to concatenation, i.e. the following
equation holds:

@@

\noindent
This means that we get the same output by applying a command $f$ to a
concatenation of two inputs $x, x'$, or by concatenating the result
produced by the application of $f$ to these inputs
separately. Similarly, it was mentioned that some pure commands (such
as \ttt{sort} and \ttt{wc}) can be parallelized using
divide-and-conquer techniques. More formally, there pure commands $f$
can be implemented as a map $m$ and a reduce $r$ function, satisfying
the following:

@@

\noindent
Similarly to the above, this means that we get the same output by
applying $f$ to a concatenation of two inputs $x, x'$, or by applying
the reduce function $r$ to the concatenation of the outputs produced
by applying map $m$ to each of these inputs.

@@

In addition to node splits, we can also perform edge splits in order
to increase the number of incoming edges for some nodes to expose
possible data parallelism. If the edge refers to a real file, then the
edge split represents splitting the real file in chunks, while in
cases where the edge connects two nodes, it is split to several edges
with a bounded number of lines. Note that the final output edge of a
valid node in the graph must always be unbounded if it does not refer
to a file.

@@

In this section, we describe challenges related to the implementation
of \sys. More concretely, in \Cref{front-end} we describe how the \sys
front-end translates a shell script to the dataflow graph that was
defined in \Cref{ir}, and in \Cref{optimizer}, we describe how the
\sys optimizer distributes and implements a given dataflow graph.

@@

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \ttt{Dish} first translates it to a
dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour, e.g. commands connected with a \ttt{\&\&}
operator. In addition, the fact that the shell is highly dynamic,
further increases the challenge of correctly identifying what parts
can be distributed, e.g. by not being able to know what file does an
unexpanded command argument refer to.

@@

In order to address this issue, we introduce the notion of
distributable program regions, and we design a translation pass that
given the abstract syntax tree of a shell script identifies
distributable regions and translates each of them to a dataflow
graph. The AST of the shell script is produced using a POSIX compliant
parser~\cite{libdash}.

\heading{Distributable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Consider the following example:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 &&
  sort f3 # Replace with sth better
\end{lstlisting}

\noindent
Notice that the structure of a shell program can be used to identify
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \ttt{cat f1 f2} and \ttt{grep
  ``foo''} would already execute concurrently in the standard shell,
while \ttt{sort f3} would wait for their completion before being
executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3}
are maximal distributable regions, in the sense that they cannot be
extended beyond \ttt{\&\&}.

Concretely, a distributable region cannot permeate through barrier
constructs (e.g. \ttt{;}, \ttt{\&\&}, and function definitions). On
the other hand, the pipe operator \ttt{|} and the parallel composition
operator \ttt{\&} enable parallelization, and extend distributable
regions. Intuitively, distributed regions correspond to subsections of
the program that would be allowed to execute concurrently by different
processes in the POSIX standard~\cite{posix}. Consider the following
illustrative example:

\tr{I don't know whether I should mention the following: While these
  control flow constructs can be parallelized and there has been
  research on it, it is orthogonal to our work, and can be
  incorporated as future work.}

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  (tr ... f1 > g1 && tr ... f2 >> g1) &
  grep ``foo'' g1
\end{lstlisting}

In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
  g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
the dataflow graph that corresponds to the whole program, has two
nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
  g1}.

\heading{Translation Pass}
%
Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom up,
translating their independent components to dataflow graph nodes,
until a barrier construct is reached. In order to identify possible
opportunities for parallelism, the translation pass also identifies
the category of each node, together with its input and output, using
records written by developers using the language that was described in
\Cref{ext}. Some interesting cases of the analysis algorithm are shown
in pseudocode below:

@@

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph (i.e. by introducing to
concurrent reads and writes to the same file when
parallelizing). Whenever the algorithm encounter an \ttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then does not merge them in a large dataflow
graph. Finally, whenever the algorithm encounters a command, it first
checks whether it is safe to parallelize it. A command is considered
to be safe, if it does not make any modification to the shell state
(e.g. changing the current directory, modifying environment
variables), and if it does not read from any read-once file
(e.g. pipe). If the command is considered safe, its category is
identified and a node containing that command is created and added to
a singleton dataflow graph. If the command is considered unsafe, it is
still added to the dataflow graph, but cannot be further parallelized
by the optimizer.

An important consideration is that due to the highly dynamic nature of
the shell, the safety properties that are mentioned above cannot be
soundly checked statically. For example in the first case, two
unexpaned strings could refer to the same file, but the translation
pass cannot infer that.

@@

The \ttt{Dish} optimizer is given a dataflow graph representing a
shell program and outputs an optimized dataflow graph, where data
parallelism opportunities have been exploited. More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges. An example
transformation can be seen in \Cref{fig:optimizer-transformation}.

@@

The optimizer also applies custom transformations on several commands
to further improve performance. Two notable examples are the
transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
that have more than one input are tranformed to \ttt{sort} commands on
each input, followed by a \ttt{sort -m}, to produce the final sorted
result. The transformation on occurences of \ttt{wc} that have more
than input \ttt{i1, t2} is illustrated below as a shell script:

@@

\heading{Merge and Split Implementation}
%
It is common in the dataflow graph for a node to have many incoming
edges, or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might only read their input from one
source (e.g. stdin). In this case, the distributed implementation
needs to concatenate input files (or conversely split output
files).
%
%% \tr{Show an example graph! of cat | tr | sort. The command in the
%%   middle must only take input from one file (maybe just stdin).}
%
We address this issue using primitive shell constructs for file
manipulation. More concretely a merger can be implemented by a simple
\verb|cat|. Given a set of input files \verb|in1, in2, ...| and an
output file \verb|out|, a merger can be implemented as:

@@

\noindent
On the other hand, given an input file \verb|in| that is a FIFO pipe,
and two output files \verb|out1, out2| the first of which is bounded
to \verb|N| lines, a 2-splitter can be implemented as:

>>>>

\subsection{Graph Transformations}
\label{ir:transformations}

Before describing the different types of parallelization
optimizations, we formalize the intuition about stateless and pure
commands that was described earlier~\sx{cmd}.
%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }
%

As mentioned there, stateless commands such as \ttt{tr} operate independently on
individual elements of the stream (characters, lines, or files)
without maintaining any state.
% Some examples include \ttt{grep}, \ttt{basename}, and \ttt{tr}.
Formally, a command $f$ is stateless if
it is ``linear'' with respect to concatenation, \ie 
% the following equation holds:
satisfying the following equation:

@@

\noindent
This means that one gets the same output by applying a command $f$ to a
concatenation of two inputs $x, x'$ as by concatenating the result
produced by the application of $f$ to these inputs
separately.

Similarly, it was mentioned that some pure commands (such
as \ttt{sort} and \ttt{wc}) can be parallelized using
divide-and-conquer techniques. More formally, these pure commands $f$
can be implemented as a combination of functions map $m$ and reduce $r$, satisfying
the following equation:

@@

\noindent
Similarly to the earlier equation, this means that we get the same output by
applying $f$ to a concatenation of two inputs $x, x'$, or by applying
the reduce function $r$ to the concatenation of the outputs produced
by applying map $m$ to each of these inputs.

@@

In addition to node splits, we can also perform edge splits in order
to increase the number of incoming edges for some nodes to further expose
possible data parallelism. In cases where the edge refers to a real file, the
edge split represents splitting the real file in chunks. In
cases where the edge connects two nodes, the split effectively refers to several edges
with a bounded number of lines. Note that the final output edge of a
valid node in the graph must always be unbounded if it does not refer
to a file.

@@

This section describes challenges related to the implementation of \sys. 
It starts by describing how the \sys front-end translates a shell script to the dataflow graph model defined in \S\ref{ir}, by applying transformation on distributable regions~\sx{front-end}.
It closes with the \sys planner, which  optimizes, distributes, and orchestrates the execution of a given dataflow graph~\sx{optimizer}.

@@

As mentioned in \Cref{ir}, in order to generate the distributed implementation of a shell script, \sys translates it to a dataflow graph.
However, there are several components in shell programs that cannot be arbitrarily parallelized without affecting the program behaviour---\eg commands connected with a \ttt{\&\&} operator~\sx{bg}.
The fact that the shell is highly dynamic compounds the challenge---\eg by not knowing what files does an unexpanded command argument refer to.
% of correctly identifying what parts can be distributed, e.g.

@@

To address this issue, we introduce the notion of distributable program regions and a translation pass for converting them to a dataflow graph.
The translation pass identifies distributable regions and translates each of them to a dataflow graph, provided the abstract syntax tree (AST) of a shell script.
The AST is produced by \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.

\paragraph{Distributable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Distributable regions are (ideally, maximal) program sub-expressions that can be parallelized safely with respect to their sequential execution.
The structure of a shell program can be used to identify components that can be executed concurrently and barriers that are natural points for enforcing synchronization.
Consider the following example:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 && sort f3 
\end{lstlisting}
% # Replace sort with sth better

\noindent
The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would already execute concurrently in the standard shell, while \ttt{sort f3} would wait for their completion before executing.
Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3} are distributable regions---maximal sub-expressions that cannot be extended beyond \ttt{\&\&}.

Intuitively, distributed regions correspond to subsections of the program that would be allowed to execute concurrently by different processes in the POSIX standard~\cite{posix}.
The pipe operator \ttt{|} and the parallel composition operator \ttt{\&} enable parallelization, and extend distributable regions.
Conversely, the \ttt{;} sequential composition operator, the \ttt{\&\&} logical operator, and function definitions represent barrier constructs that do not allow distributable regions to permeate through.

% \tr{I don't know whether I should mention the following: While these
%   control flow constructs can be parallelized and there has been
%   research on it, it is orthogonal to our work, and can be
%   incorporated as future work.}
% 
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%   (tr ... f1 > g1 && tr ... f2 >> g1) &
%   grep ``foo'' g1
% \end{lstlisting}
% 
% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
% the dataflow graph that corresponds to the whole program, has two
% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
%   g1}.
%

\paragraph{Translation Pass}
\sys performs a depth-first search pass on the AST of the given shell program.
During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.

A few illustrative cases of the analysis algorithm are shown in pseudocode below:

@@

Whenever the algorithm encounters a distributable node, it recursively
translates its subcomponents.
It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
% what first and second? we didn't say anything about first and second
Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and that at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph (\ie by introducing to
concurrent reads and writes to the same file when
parallelizing).

Whenever the algorithm encounters a barrier node,
it recursively translates its subcomponents, similarly to the distributable
case.
However, different from the distributable case, it does not merge them in a large dataflow graph.

Finally, whenever the algorithm encounters a command, it first
checks whether it is safe to parallelize it. A command is considered
safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this. % why couldn't we replicate this?
If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.

An important consideration is that due to the highly dynamic nature of
the shell, the safety properties that are mentioned above cannot be
soundly checked statically.
For safety purposes, \sys takes a conservative approach where it does not parallelize nodes without complete information.
For example, the translation pass cannot infer that two un-expand strings do not refer to the same file identifier, and as a result does not parallelize this sub-expression.

@@

Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.
More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges. An example
transformation can be seen in \Cref{fig:optimizer-transformation}.

@@

The optimizer also applies custom transformations on several commands in \pur
to further improve their performance. Two notable examples are the
transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
that have more than one input are tranformed to \ttt{sort} commands on
each input, followed by a \ttt{sort -m}, to produce the final sorted
result. The transformation on occurences of \ttt{wc} that have more
than input \ttt{i1, t2} is illustrated below as a shell script:
% \nv{do not understand i1, t2}

@@

\heading{Merge and Split Implementation}
%
It is common in the dataflow graph for a node to have many incoming
edges or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might only read their input from one
source (e.g. stdin). In this case, the distributed implementation
needs to concatenate input files (or conversely split output
files).
%
%% \tr{Show an example graph! of cat | tr | sort. The command in the
%%   middle must only take input from one file (maybe just stdin).}
%
We address this issue using primitive shell constructs for file
manipulation. More concretely a merger can be implemented by means of a simple
\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
output file \ttt{out}, a merger can be implemented as:

@@

\noindent
On the other hand, given an input file \ttt{in} that is a FIFO pipe,
and two output files \ttt{out1, out2} the first of which is bounded
to \ttt{|N} lines, a 2-splitter can be implemented as:
